---
title: "Mapping IVY-GAP Samples to Theoretical Pathway-Level Attractors w/p_Go"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r prep_data, warning=FALSE, message=FALSE}
library(tidyverse)

library(GbmPhenotypeStability)
source('../R/utils.R')
source('../R/gbmdyn.R')

# set parameters ----------------------------------------------
ivx <- GbmPhenotypeStability::ivygap$exp %>% rename(gene = gene_symbol)
iva <- GbmPhenotypeStability::ivygap$samples
pathways <- GbmPhenotypeStability::hmgsets_full
pwattr_tbl <- GbmPhenotypeStability::pwattr_tbl
# -------------------------------------------------------------
attr_pwstates <- pwattr_tbl %>%
  select(HALLMARK_E2F_TARGETS:HALLMARK_TNFA_SIGNALING_VIA_NFKB)

stopifnot(identical(iva$sample_id, colnames(ivx)[-1]))

ivx <- ivx %>% select(c('gene', iva$sample_id))
stopifnot(identical(iva$sample_id, colnames(ivx)[-1]))

xmat <- tbl_to_mat(ivx, 'gene')
stopifnot(identical(colnames(xmat), iva$sample_id))

sample_pwstates <- map_sample_exp_to_pathway_states(
  xmat,
  pathways
) %>%
mutate_all(function(x) { 
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
  }
)
rownames(sample_pwstates) <- colnames(xmat)

pwact_mat <- sample_pwstates %>% as.matrix() %>% t()
sample2loc <- purrr::set_names(iva$location, iva$sample_id)

```

---------------------------------------------------------------------------------------------------

### Ordered Pathway Activities
```{r ordered_pwact}
pltpwact_tbl <- sample_pwstates %>%
  rownames_to_column(var = 'sample_id') %>%
  gather(key = 'pathway', value = 'pathway_act', colnames(.)[-1]) %>%
  group_by(pathway) %>%
  mutate(
    rank = rank(pathway_act, ties.method = 'first')
    ) %>%
  ungroup() %>%
  arrange(pathway, rank)

shapiro_test_pval <-  function(x) { shapiro.test(x)$p.value }
dip_test_pval <- function(x) {diptest::dip.test(x)$p.value }

set.seed(1)
pwactqtl_tbl <- pltpwact_tbl %>%
  select(pathway, pathway_act) %>%
  group_by(pathway) %>%
  summarise(
    pathway_act_q50 = quantile(pathway_act, 0.50),
    pathway_act_q75 = quantile(pathway_act, 0.75),
    shapiro_test_pval = shapiro_test_pval(pathway_act),
    dip_test_pval = dip_test_pval(pathway_act)
    ) %>%
  ungroup() %>%
  arrange(pathway_act_q50)

```


# Plots to justify median-thresholding of scaled pathway score values
```{r}
plt_sample_pwstates <- sample_pwstates %>%
  rownames_to_column(var = "sample_id") %>%
  gather(key = "pathway", value = "activity", 
         HALLMARK_E2F_TARGETS:HALLMARK_TNFA_SIGNALING_VIA_NFKB) %>%
  left_join(pwactqtl_tbl, by = "pathway") %>%
  mutate(
    pathway = ifelse(pathway == "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION",
                     "HALLMARK_EMT", pathway)
  )
  
ggplot(plt_sample_pwstates, aes(x = activity)) +
  geom_histogram(aes(y=..density..), bins = 20, alpha = 0.6) + 
  geom_density(aes(y=..density..), color = "blue") +
  facet_wrap(~pathway) +
  geom_vline(aes(xintercept = pathway_act_q50), col = 'red', size = 0.75) +
  theme(strip.text = element_text(size = 8)) +
  ggtitle("Distribution of Scaled GSEA Pathway Activities for IVY-GAP Samples") +
  xlab("pathway activity") 

```



-----

### Binarizing Pathway-Level Activity Data and Matching to Attractors
+ Pathway-specific binarization (< median ---> 0, >= median ---> 1)
+ Each binary sample pathway activity vector is matched to a (binary, pathway-level)
  attractor based on the following rules:
    + Each of 'HALLMARK_E2F_TARGETS', 'HALLMARK_APOPTOSIS' must match exactly.
    + At least 2 of the GO pathways ('BIOCARTA_RAC1_PATHWAY', 'PID_CDC42_PATHWAY',   
      'BIOCARTA_RHO_PATHWAY') must match exactly.
    + Among attractors satisfying the above, the best match by Jaccard score is selected.
+ Match table is filtered to Jaccard scores > 0.5 (and satisfying the noted pathway match
  constraints, which essentially force strict(er) matching of Grow/Go/Apoptosis effector 
  pathways).
+ The Jaccard Score between binary vectors x and y is sum(x & y) / sum(x | y), where
  & and | are bitwise AND and OR, respectively.
```{r util_functions}

is_pheno_mutex <-  function(x) {
  grow      <- x[["HALLMARK_E2F_TARGETS"]]
  apoptosis <- x[["HALLMARK_APOPTOSIS"]]
  
  go_pathways <- c("BIOCARTA_RHO_PATHWAY", "PID_CDC42_PATHWAY", "BIOCARTA_RAC1_PATHWAY")
  go <- (3 == sum(x[go_pathways])) %>% as.integer()
  
  dormant <- (0 == (grow + go + apoptosis))  %>% as.integer()
  
  return(1 == (go + grow + dormant + apoptosis))
}

binarize_pwact <- function(sample_pwstates, pwact_thresholds, require_pheno_mutex = FALSE) {
  stopifnot(all(colnames(sample_pwstates) %in% names(pwact_thresholds)))
  binact <- sample_pwstates
  for (pw in colnames(binact)) {
    binact[[pw]] <- as.integer(binact[[pw]] > pwact_thresholds[[pw]])
  }
  if (!require_pheno_mutex) { return(binact) }
  
  # ---------------------------------------------------------------------------
  binarization_confidence <- function() {
    pwa <- sample_pwstates %>% as.data.frame() %>% as.matrix()
    stopifnot(!is.null(rownames(pwa)), !is.null(colnames(pwa)),
              identical(rownames(pwa), rownames(sample_pwstates)),
              identical(colnames(pwa), colnames(sample_pwstates)))
    pwa_medians <- apply(pwa, 2, median)
    pwa_mads    <- apply(pwa, 2, mad)
    stopifnot(identical(pwa_medians, pwact_thresholds[names(pwa_medians)]))
    
    # Per col: subtract column (pathway) median, divide by column MAD.
    binconf <- sweep(pwa,     MARGIN = 2, STATS = pwa_medians, FUN = "-")
    binconf <- sweep(binconf, MARGIN = 2, STATS = pwa_mads,    FUN = "/")
    # check ----------------------------------------------------
    for (i in seq_len(nrow(binconf))) {
      for (j in seq_len(ncol(binconf))) {
        conf_ij <- (pwa[[i, j]] - pwa_medians[[j]]) / pwa_mads[[j]]
        stopifnot(dplyr::near(conf_ij, binconf[[i, j]]))
      }
    }
    # ----------------------------------------------------------
    return(binconf)
  }
  
  #if (require_pheno_mutex) {
  binconf <- binarization_confidence()
  binpwa <- binact %>% as.data.frame() %>% as.matrix()
  stopifnot(identical(binconf > 0, binpwa == 1))
  
  go_pathways <- c("BIOCARTA_RHO_PATHWAY", "PID_CDC42_PATHWAY", "BIOCARTA_RAC1_PATHWAY")
  
  for (i in seq_len(nrow(binpwa))) {
    if (is_pheno_mutex(binpwa[i, , drop = TRUE])) { next }
    
    go_pw_conf_scores <- binconf[i, go_pathways, drop = TRUE]
    min_conf_go_pathway <- names(go_pw_conf_scores[which.min(go_pw_conf_scores)])
    
    pheno_conf_scores <- c(
      grow = binconf[i, "HALLMARK_E2F_TARGETS"],
      go   = min(go_pw_conf_scores),
      apoptosis = binconf[i, "HALLMARK_APOPTOSIS"]
    )
    pheno_conf_scores <- ifelse(pheno_conf_scores > 0, pheno_conf_scores, 0)
    max_conf_pheno <- names(pheno_conf_scores[which.max(pheno_conf_scores)])
    
    if (max_conf_pheno == "go") {
      stopifnot(sum(binpwa[i, go_pathways]) == 3)
      binpwa[i, "HALLMARK_E2F_TARGETS"] <- 0
      binpwa[i, "HALLMARK_APOPTOSIS"]   <- 0
    } 
    
    if (max_conf_pheno == "grow") {
      stopifnot(binpwa[i, "HALLMARK_E2F_TARGETS"] == 1)
      binpwa[i, "HALLMARK_APOPTOSIS"]   <- 0
    }
    
    if (max_conf_pheno == "apoptosis") {
      stopifnot(binpwa[i, "HALLMARK_APOPTOSIS"] == 1)
      binpwa[i, "HALLMARK_E2F_TARGETS"] <- 0
    }
    
    if ((max_conf_pheno == "grow") || (max_conf_pheno == "apoptosis")) {
      if (pheno_conf_scores[["go"]] > 0) {
        stopifnot(sum(binpwa[i, go_pathways]) == 3)
        binpwa[i, min_conf_go_pathway] <- 0
      }
    }
  }
  
  stopifnot( all(apply(binpwa, 1, function(x) { is_pheno_mutex(x) })) )
  tmp <- binpwa %>% 
    as.data.frame() %>%
    rownames_to_column(var = "sample_id") %>%
    as_tibble() %>%
    column_to_rownames(var = "sample_id")
  stopifnot(
    identical(rownames(tmp), rownames(binact)),
    identical(colnames(tmp), colnames(binact)))
  binact <- tmp

  return(binact)
}


score_mismatches <- function(x, y) {
  stopifnot(length(x) == length(y))
  result <- list()
  result[['err_count']] <- sum(x != y)
  result[['err_frac']]  <- result[['err_count']] / length(x)
  return(result)
}


score_jaccard <- function(x, y, must_match = NULL, part_match = NULL, k = 1) {
  stopifnot(length(x) == length(y), 
            all(c(x, y) %in% c(0, 1)),
            length(intersect(must_match, part_match)) == 0)
  x <- purrr::set_names(as.logical(x), names(x))
  y <- purrr::set_names(as.logical(y), names(y))
  result <- list()
  result[['jaccard_score']] <- sum(x & y) / sum(x | y)
  if (!all(x[must_match] == y[must_match])) {
    result[['jaccard_score']] <- 0
  }
  if (sum(x[part_match] == y[part_match]) < k) {
    result[['jaccard_score']] <- 0
  }
  return(result)
}

score_jaccard_broadhm = function(x, y) {
  score_jaccard(x, y, must_match = c(
    'HALLMARK_E2F_TARGETS',
    'HALLMARK_APOPTOSIS'
  ),
  part_match = c(
    'BIOCARTA_RAC1_PATHWAY',
    'PID_CDC42_PATHWAY',
    'BIOCARTA_RHO_PATHWAY'
  ),
  k = 2
  )
}


match_binary_attrs <- function(X, Y, score_func,
                               X_ids=NULL, Y_ids = NULL, use_min_score = TRUE) {
  stopifnot(identical(colnames(X), colnames(Y)))
  if (!is.null(X_ids)) {
    stopifnot(length(X_ids) == nrow(X), is.null(names(X_ids)))
  }
  if (!is.null(Y_ids)) {
    stopifnot(length(Y_ids) == nrow(Y), is.null(names(Y_ids)))
  }
  
  match_tbl <- data.frame(
    id_X = seq_len(nrow(X)),
    id_Y = NA,
    stringsAsFactors = FALSE
  )
  
  tbl_row_to_vec <- function(k, tb) {
    set_names(as.numeric(tb[k, ]), colnames(tb))
  }
  
  # Set additional columns based on (dummy) score function result.
  tmp <- score_func(tbl_row_to_vec(1, X), tbl_row_to_vec(1, X))
  for (cname in names(tmp)) { match_tbl[[cname]] <- NA }
  match_tbl <- as_tibble(match_tbl)
  
  for (i in seq_len(nrow(X))) {
    X_i <- tbl_row_to_vec(i, X)
    score_func_results  <- vector(mode = 'list', length = nrow(Y))
    
    for (j in seq_len(nrow(Y))) {
      Y_j <- tbl_row_to_vec(j, Y)
      score_func_results[[j]] <- score_func(X_i, Y_j)
    }
    
    score_vals <- map_dbl(score_func_results, function(L) L[[1]])
    if (use_min_score) {
      j_match <- which.min(score_vals)
    } else {
      j_match <- which.max(score_vals)
    }
    best_result <- score_func_results[[j_match]]
    
    match_tbl[[i, 'id_Y']] <- j_match
    for (cname in names(best_result)) {
      match_tbl[[i, cname]] <- best_result[[cname]]
    }
  }
  
  if (!is.null(X_ids)) {
    match_tbl[['id_X']] <- X_ids[match_tbl[['id_X']]]
  }
  if (!is.null(Y_ids)) {
    match_tbl[['id_Y']] <- Y_ids[match_tbl[['id_Y']]]
  }
  
  return(match_tbl)
}


get_sample_pwattr_match_tbl <- function(
  pwact_thresholds,
  sample_pwstates,
  attr_pwstates,
  score_func,
  score_colname,
  use_min_score,
  sample_meta,
  pwattr_meta,
  score_threshold = 0.5
) {
  sample_binpwact <- binarize_pwact(sample_pwstates, pwact_thresholds)
  
  match_tbl <- match_binary_attrs(
  X = sample_binpwact, X_ids = rownames(sample_binpwact),
  Y = attr_pwstates, Y_ids = rownames(attr_pwstates),
  score_func = score_func,
  use_min_score = use_min_score
  ) %>%
  rename(
    sample_id = id_X,
    pwattr_uid = id_Y
  ) %>%
  mutate(
    pwattr_uid = as.integer(pwattr_uid)
  ) %>%
  left_join(pwattr_meta, by = 'pwattr_uid') %>%
  left_join(sample_meta, by = 'sample_id')
  
  if (use_min_score) {
    selector <- match_tbl[[score_colname]] < score_threshold
  } else {
    selector <- match_tbl[[score_colname]] > score_threshold
  }
  
  match_tbl %>% filter(selector)
}
```


```{r attr_matching}
median_thresholds <- deframe(pwactqtl_tbl %>% select(pathway, pathway_act_q50))

match_tbl <- get_sample_pwattr_match_tbl(
  pwact_thresholds = median_thresholds,
  sample_pwstates = sample_pwstates,
  attr_pwstates = attr_pwstates,
  score_func = score_jaccard_broadhm,
  score_colname = 'jaccard_score',
  use_min_score = FALSE,
  sample_meta = iva %>% select(sample_id, location, stem_cluster_status),
  pwattr_meta = pwattr_tbl %>% select(pwattr_uid, p_Go),
  score_threshold = 0.5
) %>%
  left_join(select(pwattr_tbl, pwattr_uid, p_Grow, p_Dormant, p_Apoptosis),
            by = "pwattr_uid") %>%
  select(sample_id, pwattr_uid, jaccard_score, p_Go, p_Grow, p_Dormant, p_Apoptosis,
         everything())

stopifnot(select(match_tbl, p_Go:p_Apoptosis) %>%
            mutate(p_All = p_Go + p_Grow + p_Dormant + p_Apoptosis) %>%
            pull(p_All) %>% 
            near(1) %>%
            all())

stopifnot(identical(match_tbl, GbmPhenotypeStability::ivygap$pwattr_match_tbl))
# ivygap <- GbmPhenotypeStability::ivygap
# ivygap$pwattr_match_tbl <- match_tbl
# save(ivygap, file = "data/ivygap.RData")
```

-------------------------------------------------------------------------------

### Write Out Supplementary Table

```{r}
ivygap_pwattr_pGo_tbl <- GbmPhenotypeStability::ivygap$pwattr_match_tbl %>%
  left_join(pwattr_tbl %>% select(pwattr_uid, phenotype,
                                  HALLMARK_E2F_TARGETS:HALLMARK_TNFA_SIGNALING_VIA_NFKB), 
            by = "pwattr_uid") %>%
  rename(
    ivygap_sample_id = sample_id, 
    pwattr_phenotype = phenotype
    ) %>%
  mutate(
    pwattr_phenotype = purrr::map_chr(pwattr_phenotype, function(s) { paste0(s, collapse = ", ") })
    )%>%
  select(ivygap_sample_id, pwattr_uid, pwattr_phenotype, everything()) %>%
  arrange(p_Go)

# CHECK -----------------------------------------------------------------------
# tmp <- readxl::read_xlsx(
#   "~/Dropbox/manuscripts/gbm_manuscript/Submission_Version/Table_S5_ivygap_pwattractor_pGo_table.xlsx") %>%
#   mutate(
#     ivygap_sample_id = as.character(ivygap_sample_id),
#     pwattr_uid = as.integer(pwattr_uid)
#     ) %>%
#   arrange(p_Go)
# 
# stopifnot(
#   identical(
#     ivygap_pwattr_pGo_tbl %>% select(-jaccard_score, -p_Go, -p_Grow, -p_Dormant, -p_Apoptosis),
#     tmp %>% select(-jaccard_score, -p_Go)
#   ),
#   all(near(ivygap_pwattr_pGo_tbl$p_Go, tmp$p_Go)),
#   all(near(ivygap_pwattr_pGo_tbl$jaccard_score, tmp$jaccard_score))
# )
# -----------------------------------------------------------------------------

# writexl::write_xlsx(
#   ivygap_pwattr_pGo_tbl, 
#   "inst/extdata/Sup_Tables/Table_S5_ivygap_pwattractor_pGo_table.xlsx")

```

